Supplements for ``Tensor denoising and completion based on \\ ordinal observations''

### 2020-06-23 paper/Camera ready/supp.tex
Add files via upload

\newcommand*{\KeepStyleUnderBrace}[1]{%f
  \mathop{%
    \mathchoice
    {\underbrace{\displaystyle#1}}%
    {\underbrace{\textstyle#1}}%
    {\underbrace{\scriptstyle#1}}%
    {\underbrace{\scriptscriptstyle#1}}%
  }\limits
}
\usepackage{makecell}
\input macros.tex

\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

\twocolumn[
\icmltitle{Tensor denoising and completion based on ordinal observations}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2020
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
%\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Chanwoo Lee}{wi}
\icmlauthor{Miaoyan Wang}{wi}
\end{icmlauthorlist}

\icmlaffiliation{wi}{Department of Statistics, University of Wisconsin-Madison, Wisconsin, USA}
\icmlcorrespondingauthor{Miaoyan Wang}{miaoyan.wang@wisc.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Higher-order tensors, ordinal observation, tensor decomposition, tensor completion}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
Higher-order tensors arise frequently in applications such as neuroimaging, recommendation system, and social network analysis. We consider the problem of low-rank tensor estimation from possibly incomplete, ordinal-valued observations. Two related problems are studied, one on tensor denoising and another on tensor completion. We propose a multi-linear cumulative link model, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. Our mean squared error bound enjoys a faster convergence rate than previous results, and we show that the proposed estimator is minimax optimal under the class of low-rank models. Furthermore, the procedure developed serves as an efficient completion method which guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tilde \tO(Kd)$ noisy, quantized observations. We demonstrate the outperformance of our approach over previous methods on the tasks of clustering and collaborative filtering.

\end{abstract}

\section{Introduction}
Multidimensional arrays, a.k.a.\ tensors, arise in a variety of applications including recommendation systems~\citep{baltrunas2011incarmusic}, social networks~\citep{nickel2011three}, genomics~\citep{wang2019three}, and neuroimaging~\citep{zhou2013tensor}. There is a growing need to develop general methods that can address two main problems for analyzing these noisy, high-dimensional datasets. The first problem is tensor denoising which aims to recover a signal tensor from its noisy entries~\citep{hong2020generalized,wang2019multiway}. The second problem is tensor completion which examines the minimum number of entries needed for a consistent recovery~\citep{ghadermarzy2018learning,montanari2018spectral}. Low-rankness is often imposed to the signal tensor, thereby efficiently reducing the intrinsic dimension in both problems.

A number of low-rank tensor estimation methods have been proposed~\citep{hong2020generalized,wang2017tensor}, revitalizing classical methods such as CANDECOMP/PARAFAC (CP) decomposition~\citep{hitchcock1927expression} and Tucker decomposition~\citep{tucker1966some}. These tensor methods treat the entries as continuous-valued. In many cases, however, we encounter datasets of which the entries are qualitative. For example, the Netflix problem records the ratings of users on movies over time. Each data entry is a rating on a nominal scale \{{\it very like, like, neutral, dislike, very dislike}\}. Another example is in the signal processing, where the digits are frequently rounded or truncated so that only integer values are available. The qualitative observations take values in a limited set of categories, making the learning problem harder compared to  continuous observations.

Ordinal entries are categorical variables with an ordering among the categories; for example, {\it very like} $\prec$ {\it like} $\prec$ {\it neutral} $\prec \cdots$. The analyses of tensors with the ordinal entries are mainly complicated by two key properties needed for a reasonable model. First, the model should be invariant under a reversal of categories, say, from the Netflix example, {\it very like} $\succ$ {\it like} $\succ$ {\it neutral} $\succ \cdots$, but not under arbitrary label permutations. Second, the parameter interpretations should be consistent under merging or splitting of contiguous categories. The classical continuous tensor model~\cite{kolda2009tensor, ghadermarzy2019near} fails in the first aspect, whereas the binary tensor model~\cite{ghadermarzy2018learning} lacks the second property. An appropriate model for ordinal tensors has yet to be studied.

\begin{table*}[ht]
\centerline{\includegraphics[width=1\textwidth]{compare.pdf}}
\caption{Comparison with previous work. For ease of presentation, we summarize the error rate and sample complexity assuming equal tensor dimension in all modes. $K$: tensor order; $L$: number of ordinal levels; $d$: dimension at each mode. }\label{tab:compare}
\end{table*}

{\bf Our contributions.} We are among the first to establish the recovery theory for signal tensors and quantization operators simultaneously from a limited number of highly discrete entries. Our main contributions are summarized in Table~\ref{tab:compare}. We propose a cumulative link model for higher-order tensors, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. The mean squared error bound is established, and we show that the obtained bound has minimax optimal rate in high dimensions under the low-rank model. Furthermore, our proposal guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tilde \tO(Kd)$ noisy, quantized observations.

{\bf Related work.} Our work is connected to non-Gaussian tensor decomposition. Existing work focuses exclusively on univariate observations such as binary- or continuous-valued entries~\citep{wang2018learning,ghadermarzy2018learning}.  The problem of ordinal/quantized tensor is fundamentally more challenging than previously well-studied tensors for two reasons: (a) the entries do not belong to exponential family distribution, and (b) the observation contains much less information, because neither the underlying signal nor the quantization operator is known. The limited information makes the statistical recovery notably hard.

We address the challenge by proposing a cumulative link model that enjoys palindromic invariance~\cite{mccullagh1980regression}. A distinctive non-monotonic, phase-transition pattern is demonstrated (Fig~\ref{fig:finite}b). We prove that recovery from quantized tensors achieves equally good information-theoretical convergence as the continuous tensors. These results fills the gap between classical and non-classical (ordinal) observations, thereby greatly enriching the tensor-model literature.

From algorithm perspective, we address the challenge using the (non-convex) alternating algorithm. Earlier work has proposed an approximate (convex) algorithm for binary tensor completion~\cite{ghadermarzy2018learning}. Unlike matrix problems, convex-relaxation for low-rank tensors suffers from both computational intractability~\cite{hillar2013most} and statistical suboptimality. We improve the error bound from $\tO(d^{-(K-1)/2})$ in~\citet{ghadermarzy2018learning} to $\tO(d^{-(K-1)})$ and numerically compare the two approaches.

We also highlight the challenge associated with higher-order tensors. Matrix completion has been proposed for binary observations~\cite{cai2013max,davenport2014,bhaskar20151} and for ordinal observations~\cite{bhaskar2016probabilistic}. We show that, applying existing matrix methods to higher-order tensors results in suboptimal estimates. A full exploitation of the higher-order structure is needed; this is another challenge we address in this paper.

\section{Preliminaries}
Let $\tY\in\mathbb{R}^{d_1\times \cdots \times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. We use $y_\omega$ to denote the tensor entry indexed by $\omega$, where $\omega\in[d_1]\times\cdots\times[d_K]$.  The Frobenius norm of $\tY$ is defined as $\FnormSize{}{\tY}=\sum_\omega y^2_\omega$ and the infinity norm is defined as $\mnormSize{}{\tY}=\max_{\omega}|y_\omega|$. We use $\tY_{(k)}$ to denote the unfolded matrix of size $d_k$-by-$\prod_{i\neq k}d_k$, obtained by reshaping the tensor along the mode-$k$. The Tucker rank of $\tY$ is defined as a length-$K$ vector $\mr = (r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\tY_{(k)}$ for $k \in[K]$.

We use lower-case letters ($a, b, \ldots$) for scalars/vectors, upper-case boldface letters ($\mA, \mB,  \ldots$) for matrices, and calligraphy letters ($\tA,\tB,\ldots$) for tensors of order three or greater.  An event $A$ is said to occur ``with very high probability'' if $\mathbb{P}(A)$ tends to 1 faster than any polynomial of tensor dimension $d_{\min}=\min\{d_1,\ldots,d_K\} \to\infty$. The indicator function of an event $A$ is denoted as $\mathds{1}\{A\}$. For ease of notation, we allow basic arithmetic operators (e.g., $\leq, +, -$) to be applied to pairs of tensors in an element-wise manner. We use the shorthand $[n]$ to denote $\{1,\ldots,n\}$ for $n \in N_{+}$.

\section{Model formulation and motivation}
\subsection{Observation model}
Let $\tY$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional data tensor. Suppose the entries of $\tY$ are ordinal-valued, and the observation space consists of $L$ ordered levels, denoted by $[L]:=\{1,\ldots,L\}$. We propose a cumulative link model for the ordinal tensor $\tY=\entry{y_\omega}\in[L]^{d_1\times \cdots\times d_K}$. Specifically, assume the entries $y_\omega$ are (conditionally) independently distributed with cumulative probabilities:
\begin{equation}\label{eq:model}
\mathbb{P}(y_\omega\leq \ell)=f(b_\ell-\theta_\omega),\ \text{for all}\ \ell\in[L-1],
\end{equation}
where $\mb=(b_1,\ldots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots <b_{L-1}$, $\Theta=\entry{\theta_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a continuous-valued parameter tensor satisfying certain low-dimensional structure (to be specified later), and $f(\cdot)\colon \mathbb{R}\mapsto[0,1]$ is a known, strictly increasing function. We refer to $\mb$ as the cut-off points and $f$ the link function.

The formulation~\eqref{eq:model} imposes an additive model to the transformed probability of cumulative categories. This modeling choice is to respect the ordering structure among the categories. For example, if we choose the inverse link $f^{-1}(x)=\log {x\over 1-x}$ to be the log odds, then the model~\eqref{eq:model} implies linear spacing between the proportional odds:
\begin{equation}\label{eq:logodd}
\log {\mathbb{P}(y_{\omega}\leq \ell) \over \mathbb{P}(y_\omega >  \ell) } - \log {\mathbb{P}(y_{\omega}\leq {\ell-1}) \over \mathbb{P}(y_\omega >  {\ell-1}) } = b_\ell-b_{\ell-1},
\end{equation}
for all tensor entries $y_\omega$. When there are only two categories in the observation space (e.g. binary tensors), the cumulative model~\eqref{eq:model} is equivalent to the usual binomial link model. In general, however, when the number of categories $L\geq 3$, the proportional odds assumption~\eqref{eq:logodd} is more parsimonious, in that, the ordered categories can be envisaged as contiguous intervals on the continuous scale, where the points of division are exactly $b_1<\cdots <b_{L-1}$. This interpretation will be made more explicit in the next section.

\subsection{Latent-variable interpretation}\label{sec:latent}

The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
%y_\omega=\sum_{\ell\in[L]} \ell \mathds{1}\{y^*_\omega \in(b_{\ell-1},b_\ell]\},\ \forall \ \omega\in[d_1]\times \cdots \times [d_k],
%\end{equation}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

The equivalence between the latent-variable model~\eqref{eq:quantization} and the cumulative link model~\eqref{eq:model} is established if the link $f$ is chosen to be the cumulative distribution function of noise $\varepsilon$, i.e., $f(\theta)=\mathbb{P}(\varepsilon\leq \theta)$. We describe two common choices of link $f$, or equivalently, the distribution of $\varepsilon$.

\begin{example}[Logistic model] The logistic model is characterized by~\eqref{eq:model} with $f(\theta)=(1+e^{-\theta/\sigma})^{-1}$, where $\sigma>0$ is the scale parameter. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ logistic distribution with scale parameter $\sigma$.
\end{example}
\begin{example}[Probit model] The probit model is characterized by~\eqref{eq:model} with
$f(\theta)=\mathbb{P}(z\leq \theta/\sigma)$, where $z\sim N(0,1)$. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ $N(0,\sigma^2)$.
\end{example}
Other link functions are also possible, such as Laplace, Cauchy, etc~\cite{mccullagh1980regression}. All the models share the property that the ordered categories can be thought of as contiguous interval on some continuous scale. We should point out that, although the latent-variable interpretation is incisive, our estimation procedure does not refer to the existence of $\tY^*$. Therefore, our model~\eqref{eq:model} is general and still valid in the absence of quantization process. More generally, we make the following assumptions about the link $f$.

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.2cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing and twice-differentiable in $\theta\in \mathbb{R}$.
\item $f'(\theta)$ is strictly log-concave and symmetric with respect to $\theta=0$.
\end{enumerate}
\end{assumption}

\subsection{Problem 1: Tensor denoising}~\label{sec:denoising}
The first question we aim to address is tensor denoising:

(P1) Given the quantization process induced by $f$ and the cut-off points $\mb$, how accurately can we estimate the latent signal tensor $\Theta$ from the ordinal observation $\tY$?

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space:
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor data analysis, and is shown to provide a reasonable tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notations of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notation of low-rankness may lead to mathematically ill-posed optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well-posedness of optimization and easy interpretation.

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The entry-wise bound assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.

\subsection{Problem 2: Tensor completion}
Motivated by applications in collaborative filtering, we also consider a more general setup when only a subset of tensor entries $y_\omega$ are observed. Let $\Omega\subset[d_1]\times \cdots\times[d_K]$ denote the set of observed indices. The second question we aim to address is stated as follows:

(P2) Given an incomplete set of ordinal observations $\{y_{\omega}\}_{\omega\in\Omega}$, how many sampled entries do we need to consistently recover $\Theta$ based on the model~\eqref{eq:model}?

The answer to (P2) depends on the choice of $\Omega$. We consider a general model on $\Omega$ that allows both uniform and non-uniform sampling. Specifically, let $\Pi=\{\pi_{i_1,\ldots,i_K}\}$ denote a predefine probability distribution over the index set such that $\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_\omega =1$. We assume that each index in $\Omega$ is drawn with replacement using distribution $\Pi$. This sampling model relaxes the uniform sampling in literature and is arguably a better fit in applications.

We consider the same parameter space~\eqref{eq:space} for the completion problem. In addition to the reasons mentioned in Section~\ref{sec:denoising}, the entrywise bound assumption also serves as the incoherence requirement for completion. In classical matrix completion, the incoherence is often imposed on the singular vectors. This assumption is recently relaxed for ``flat'' matrices with bounded magnitude~\cite{negahban2011estimation,cai2013max,bhaskar20151}. We adopt the same assumption for higher-order tensors.

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}\{y_\omega=\ell\} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. We will focus on the theory and algorithm with known $\mb$ in the main paper. The adaptation for unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {|\dot{g}_\ell(\theta)|\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry including flatness and convexity of the latent noise distribution. Under the Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimension.

\subsection{Estimation error for tensor denoising}\label{sec:denosing}
For the tensor denoising problem, we assume that the full set of tensor entries are observed. We assess the estimation accuracy using the mean squared error (MSE):
\[
\text{MSE}(\hat \Theta, \trueT)={1\over \prod_k d_k}\FnormSize{}{\Theta-\trueT}^2.
\]
The next theorem establishes the upper bound for the MSE of the proposed $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\mathrm{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_1  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1 >0$ is a constant that depends only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

A similar conclusion is obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the categorical distributions in the observation space.
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability,
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_1 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1 >0$ is the same constant as in Theorem~\ref{thm:rate}.
\end{cor}
To gain insight into these bounds, we consider a special setting with equal dimension in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation}
Hence, our estimator achieves consistency with polynomial convergence rate. We compare the bound with existing literature. In the special case $L=2$, \citet{ghadermarzy2018learning} proposed a max-norm constrained estimator $\tilde{\Theta}$ with $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$. In contrast, our estimator converges at a rate of $d^{-(K-1)}$, which is substantially faster than theirs. This provides a positive answer to the open question posed in~\citet{ghadermarzy2018learning} whether the square root in the bound is removable. The improvement stems from utilizing the exact low-rankness of $\Theta$, whereas the surrogate rank measure employed in~\citet{ghadermarzy2018learning} is scale-sensitive.

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix estimation was $\tO(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorable as the order of data tensor increases. Intuitively, the sample size for tensor data analysis is the number of entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=\tO(1)$. A higher tensor order implies higher effective sample size per parameter, and thus exhibits a faster convergence rate in high dimensions.

We next show the statistical optimality of our estimator $\hat \Theta$. The result is based on the information theory, and applies to all estimators in $\tP$, including but not limited to $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Minimax lower bound]\label{thm:minimax}
Assume the same set-up as in Theorem~\ref{thm:rate}, and $d_{\max}=\max_k d_k \geq 8$. Let $\inf_{\hat \Theta}$ denote the infimum over all estimators $\hat \Theta\in\tP$ based on the ordinal tensor observation $\tY\in[L]^{d_1\times \cdots \times d_K}$. Then, under the model~\eqref{eq:model},
\begin{align}\label{eq:lower}
\inf_{\hat \Theta }\sup_{\trueT\in\tP} \mathbb{P}\Big\{& \text{MSE}(\hat \Theta, \trueT) \\
& \geq c\min\left( \alpha^2, \ { Cr_{\max}d_{\max} \over \prod_k d_k} \right) \Big\} \geq {1\over 8},
\end{align}
where $C=C(\alpha, L, f,\mb)>0$ and $c>0$ are constants independent of tensor dimension and the rank.
\end{thm}
We see that the lower bound matches the upper bound in~\eqref{eq:rate} on the polynomial order of tensor dimension. Therefore, our estimator~\eqref{eq:estimator} is rate-optimal.

\subsection{Sample complexity for tensor completion}
We now consider the tensor completion problem, when only a subset of entries $\Omega$ are observed. We consider a general sampling procedure induced by $\Pi$. The recovery accuracy is assessed by the weighted squared error:
\begin{align}\label{eq:weighted}
\PiFnormSize{}{\Theta-\hat \Theta}^2&\stackrel{\text{def}}{=}
{1\over |\Omega|}\mathbb{E}_{\Omega\sim \Pi}\FnormSize{}{\Theta-\hat \Theta}^2\notag \\
&=\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_{\omega}(\Theta_{\omega}-\hat \Theta_{\omega})^2.
\end{align}

Note that the recovery error depends on the distribution $\Pi$. In particular, tensor entries with higher sampling probabilities have more influence on the recovery accuracy, compared to the ones with lower sampling probabilities.
\begin{rmk} If we assume each entry is sampled with strictly positive probability; i.e.\ there exits a constant $\mu> 0$ s.t.
\[
\ \pi_\omega\geq {1\over \mu \prod_k d_k},\quad \text{for all}\ \omega\in[d_1]\times\cdots \times [d_K],
\]
then the error in~\eqref{eq:weighted} provides an upper bound for MSE:
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \geq {\FnormSize{}{\Theta-\hat \Theta}^2\over \mu \prod_kd_k}={1\over \mu}\text{MSE}(\hat \Theta, \trueT).
\]
The equality is attained under uniform sampling with $\mu=1$.
\end{rmk}

\begin{thm}~\label{thm:completion}
Assume the same set-up as in Theorem~\ref{thm:rate}. Suppose that we observe a subset of tensor entries $\{y_\omega\}_{\omega\in\Omega}$, where $\Omega$ is chosen at random with replacement according to a probability distribution $\Pi$. Let $\hat \Theta$ be the solution to~\eqref{eq:estimator}, and assume $r_{\max}=\tO(1)$. Then, with very high probability,
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \to 0,\quad \text{ as }\quad {|\Omega|\over \sum_k d_k} \ \to \infty.
\]
\end{thm}
Theorem~\ref{thm:completion} shows that our estimator achieves consistent recovery using as few as $\tilde{\tO}(Kd)$ noisy, quantized observations from an order-$K$ $(d,\ldots,d)$-dimensional tensor. Note that $\tilde \tO(Kd)$ roughly matches the degree of freedom for an order-$K$ tensor of fixed rank $\mr$, suggesting the optimality of our sample requirement. This sample complexity substantially improves over earlier result $\tO(d^{\lceil K/2\rceil})$ based on square matricization~\cite{mu2014square}, or $\tO(d^{K/2})$ based on tensor nuclear-norm regularization~\cite{yuan2016tensor}. Existing methods that achieve $\tilde \tO(Kd)$ sample complexity require either a deterministic cross sampling design~\cite{zhang2019cross} or univariate measurements~\cite{ghadermarzy2018learning}. Our method extends the conclusions to multi-level measurements under a broader class of sampling schemes.

\section{Numerical Implementation}\label{sec:algorithm}
We describe the algorithm to seek the optimizer of~\eqref{eq:objective}. In practice, the cut-off points $\mb$ are often unknown, so we choose to maximize $\logl$ jointly over $(\Theta,\mb)\in\tP\times \tB$ (see the Supplement for details). The objective $\logl$ is concave in $(\Theta,\mb)$ whenever $f'$ is log-concave. However, the feasible set $\tP$ is non-convex, which makes the optimization~\eqref{eq:objective} a non-convex problem. We employ the alternating optimization approach by utilizing the Tucker representation of $\Theta$. Specifically, based on~\eqref{eq:Tucker} and~\eqref{eq:objective}, the objective function consists of $K+2$ blocks of variables, one for the cut-off points $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. The optimization is a simple convex problem if any $K+1$ out of the $K+2$ blocks are fixed. We update one block at a time while holding others fixed, and alternate the optimization throughout the iteration. The convergence is guaranteed whenever $\logl$ is bounded from above, since the alternating procedure monotonically increases the objective. The Algorithm~\ref{alg} gives the full description.

>>>>

\newcommand*{\KeepStyleUnderBrace}[1]{%f
  \mathop{%
    \mathchoice
    {\underbrace{\displaystyle#1}}%
    {\underbrace{\textstyle#1}}%
    {\underbrace{\scriptstyle#1}}%
    {\underbrace{\scriptscriptstyle#1}}%
  }\limits
}
\usepackage{makecell}
\input macros.tex
\usepackage{xr}
\externaldocument{ordinalTv6}
\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

\appendix
\onecolumn
